---
title: "Analysis update"
subtitle: "Analysis of scRNA-Seq data from Bennet"
author: "Maria Schmidt"
affiliation: "IZBI, Interdisciplinary Centre for Bioinformatics, Leipzig University, Germany"
date: "`r Sys.Date()`"
format:
  html: 
    toc: true
    number-sections: false
    code-fold: true
  pptx: default
  docx: default
link-citations: true
params:
  run: "Reynolds2021_Data-all_seuratObj"
---


```{r setup}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

```


```{r}
pacman::p_load(Seurat, tidyverse, here, ggplot2, pals, flextable, patchwork)
source(here("src","seurat-functions.r"))
source(here("src","make-nicknames-for-celltypes.r"))
output_data_path <- "output"
input_data_path <- "data"

if(params$run == "Reynolds2021_Data-all_seuratObj" ){
  
      load(here("output", paste0(params$run,".RData")))
      cell_type = "Cell_type_nicknames"
      orig_ident <- "Flow_gate"
      unfiltered_matrix_ncol <- ncol(seurat_object)
      seurat_object$Cell_type_nicknames <- make_nicknames(seurat_object$Cell_type)
      celltype_colors <- randomcoloR::distinctColorPalette(length(table(seurat_object@meta.data[, cell_type])))
      celltype_cluster_data <- data.frame(
          table(seurat_object@meta.data[, cell_type]),
          "color" = celltype_colors )
    
    } else {
      
      seurat_object <- load_seurat_object(data_set_name = params$run,  output_data_path = output_data_path)
      cell_type = "predicted.id.Cell_type_nicknames"
      orig_ident = "orig.ident"
      unfiltered_matrix_ncol <- ncol(Read10X(data.dir =  here(input_data_path, params$run, "filtered_feature_bc_matrix")))
      # seurat_object$predicted.id.Cell_type_nicknames <- make_nicknames(seurat_object$predicted.id.Cell_type)
      celltype_colors <- randomcoloR::distinctColorPalette(length(unique(seurat_object@meta.data[, cell_type])))
      names(celltype_colors) <- sort(unique(seurat_object@meta.data[, cell_type]))
      celltype_cluster_data <- data.frame(
          table(seurat_object@meta.data[, cell_type])[which(table(seurat_object@meta.data[, cell_type])>0)],
          "color" = celltype_colors
      )
}
seurat_colors <- scales::hue_pal()(length(unique(seurat_object@meta.data[, "seurat_clusters"])))
seurat_cluster_data <- data.frame(
    table(seurat_object@meta.data[, "seurat_clusters"]),
    "color" = seurat_colors
)
filename <- here(output_data_path, "bennets-skin-marker.RData")
load(filename)
filename <- here(output_data_path, "reynolds-skin-marker.RData")
load(filename) 
filename <- here(output_data_path, "interleukins.RData")
load(filename)
# filename <- here(output_data_path, "TXLNA-cells-markers.RData")
# load(filename)
# filename <- here(output_data_path, "IFNL1-cells-markers.RData")
# load(filename)
filename <- here(output_data_path, paste( params$run, cell_type, "marker.RData", sep = "_"))
load(filename)


```

# Preprocessing, data integration and dimension reduction

## Quality check

```{r, echo=FALSE,  fig.width = 16, fig.height = 11}
p1 <- VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = cell_type, cols = celltype_colors)
p2 <- VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = "seurat_clusters")

p1 / p2
```

```{r, echo=FALSE,  fig.width = 7, fig.height = 5}
VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = orig_ident)
FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "percent.mt", group.by = orig_ident)
ggplot(seurat_object@meta.data) +
    geom_point(aes(x = nCount_RNA, y = nFeature_RNA, colour = percent.mt)) +
    theme_minimal()
```


```{r, echo=FALSE,  fig.width = 7, fig.height = 7}
ft <- continuous_summary(
  seurat_object@meta.data,
  columns = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rps"),
  by = orig_ident,
  hide_grouplabel = TRUE,
  digits = 3
)
ft <-  set_caption(ft, caption = "Statistics of number of counts, features, mitochondrial expression and ribosomal expression")
ft
```

## Cell-cycle scoring

```{r, echo=FALSE,  fig.width = 6, fig.height = 5}
DimPlot(seurat_object, reduction = "umap", label = FALSE, group.by = "Phase")
```

## Unsupervised louvain clustering

To cluster cell in an unsupervised fashion, we applied the Louvain-algorithm implemented in the Seurat package.

```{r, echo=FALSE,  fig.width = 10, fig.height = 5}
p1 <- DimPlot(seurat_object, reduction = "umap", label = TRUE, repel = TRUE, group.by = "seurat_clusters") + theme(legend.position = "none")
p2 <- ggplot(seurat_cluster_data, aes(x = Var1, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  scale_fill_manual(values = seurat_cluster_data[, "color"]) +
  coord_flip() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  ) +
  xlab("Cluster") +
  ylab("Frequency")

p1 + p2
```

## Celltype classification using a reference dataset

We projected celltypes from a healthy scRNA-Seq dataset (Reynolds et al., 2021) onto our dataset.

```{r, echo=FALSE,  fig.width = 10, fig.height = 5}

Idents(seurat_object) <- cell_type

p1 <- DimPlot(seurat_object, reduction = "umap", group.by = cell_type, cols = celltype_colors) + theme(legend.position = "none", legend.text = element_text(size = 7)) + ggtitle("")

p2 <- ggplot(celltype_cluster_data, aes(x = Var1, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  scale_fill_manual(values = celltype_colors) +
  coord_flip() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  ) +
  xlab("Celltype") +
  ylab("Frequency")

p1 + p2
```

To verify this further, we examined some canonical cell type markers for the specific skin and immune cell populations.

```{r, echo=FALSE, fig.align = "center", fig.height = 10, fig.width = 15 }
DotPlot(seurat_object, features = skin_marker, group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 10)
  )
```

We calculated positive markers via differential expression, which further verify and define our predicted clusters .

```{r, echo=FALSE, fig.align = "center", fig.height = 20, fig.width = 20 }

Idents(seurat_object) <- cell_type
# cluster_marker <- FindAllMarkers(seurat_object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
# save(cluster_marker, file = here(output_data_path, paste( params$run, cell_type, "marker.RData", sep = "_")))
top5_predicted.id.Cell_type <- cluster_marker %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC)
DoHeatmap(seurat_object, features = top5_predicted.id.Cell_type$gene, group.by = cell_type, group.bar.height = 0.03, group.colors = celltype_cluster_data$color, hjust = 0) +
  xlab("Celltype") +
  ylab("Top 5 Marker per Cluster") + theme(legend.position = "none") +
  theme(
    axis.text.y = element_text(size = 15),
    axis.title = element_text(size = 30),
    axis.text.x = element_text(size = 25)
  ) +
  scale_fill_gradient2(high = "red", low = "blue")
```


# Interleukine (IL) expression

## All differentilly expressed ILs

```{r echo=FALSE, fig.align="center", fig.height=6, fig.width=9}
interleukins <- intersect(interleukins, rownames(seurat_object))
interleukins <- unique(c(interleukins, rownames(seurat_object)[grep("^IL", rownames(seurat_object))]))
interleukins.diffexpr <- unique(cluster_marker$gene[which(cluster_marker$gene %in% interleukins)])

DotPlot(seurat_object, features = interleukins.diffexpr, group.by = cell_type, dot.scale = 5) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13)
  )
```

## All differentilly expressed ILs in inflammatory celltypes

```{r, echo=FALSE, fig.align = "center", fig.height = 5, fig.width = 5 }
#  (also die DC- und die T-Fraktion, dann gerne Th1/Th2-aufgeschlüsselt, Makrophagen, Monocyten, Plasmazellen, Mastzellen)
dendritic <- grep("DC", unique(seurat_object@meta.data[, cell_type]), value = TRUE)
tcells <- c("Tc", "Th", "Treg")
macrophages <- grep("Mac", unique(seurat_object@meta.data[, cell_type]), value = TRUE)
langerhans <- grep("LC", unique(seurat_object@meta.data[, cell_type]), value = TRUE)
inflammatory <- sort(c(dendritic, tcells, macrophages, langerhans, "Mast cell", "Monocyte", "Plasma cell" ))
interleukins_inflammatory_diffexpr <- unique(cluster_marker$gene[which(cluster_marker$gene %in% interleukins & cluster_marker$cluster %in% inflammatory)])

Idents(seurat_object) <- cell_type

```


```{r, echo=FALSE, fig.align = "center", fig.height = 5, fig.width = 13}

DotPlot(seurat_object, features = interleukins_inflammatory_diffexpr, group.by = cell_type, dot.scale = 5, idents = inflammatory) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13)

  )
```


## All expressed ILs

```{r, echo=FALSE, fig.align = "center", fig.height = 8, fig.width = 20 }

DotPlot(seurat_object, features = interleukins, group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
    axis.text.y = element_text(size = 20)
  )
```

## All expressed ILs in inflammatory celltypes

```{r, echo=FALSE, fig.align = "center", fig.height = 6, fig.width = 20 }

DotPlot(seurat_object, features = interleukins, group.by = cell_type, dot.scale = 6, idents = inflammatory) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
    axis.text.y = element_text(size = 20)
  )
```

## Selected ILs/Features

```{r, echo=FALSE, fig.align = "center",  fig.height = 50, fig.width = 12 }
genes.bennet <- intersect(c(rownames(seurat_object)[grep("IL17", rownames(seurat_object))], "TNF", "TNFRSF1A", "IL22", rownames(seurat_object)[grep("IL12", rownames(seurat_object))], "IL23R", "IL23A", "NOS1", "NOS2", "IL13", "IL13RA1"), rownames(seurat_object))
Idents(seurat_object) <- cell_type
FeaturePlot(
  seurat_object, 
  features = genes.bennet,
  ncol = 2,
  order = TRUE, 
  label = TRUE, 
  repel = TRUE
  )
```

```{r, echo=FALSE, fig.align = "center", fig.height = 7, fig.width = 10}
DotPlot(seurat_object, features = genes.bennet, group.by = cell_type, dot.scale = 8) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12)
  )

```

## Selected ILs/Features in inflammatory celltypes

```{r, echo=FALSE, fig.align = "center",  fig.height = 5, fig.width = 12 }
DotPlot(seurat_object, features = genes.bennet, group.by = cell_type, dot.scale = 8, idents = inflammatory) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 14)
  )

```


# Interesting interferons (IFNs)

## Interferone (alpha, beta, gamma), TNF-alpha, TGFalpha

```{r, echo=FALSE, fig.align = "center", fig.height = 8, fig.width = 6 }
DotPlot(seurat_object, features = c("TNF", "LTA",  "TGFA", "TGFB1"), group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
```

```{r, echo=FALSE, fig.align = "center", fig.height = 8, fig.width = 11 }
FeaturePlot(
  seurat_object, 
  features = c("TNF", "LTA",  "TGFA", "TGFB1"),
  ncol = 2,
  order = TRUE, 
  label = TRUE, 
  repel = TRUE, 
  min.cutoff = "q1",
  max.cutoff = "q99"
  )

```


## All TNFs and TGFs

```{r, echo=FALSE, fig.align = "center", fig.height = 8, fig.width = 15}
interferons <- c(rownames(seurat_object)[grep("^TNF", rownames(seurat_object))], rownames(seurat_object)[grep("^TGF", rownames(seurat_object))])

DotPlot(seurat_object, features = interferons, group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
```


# S100A1 proteins

```{r, echo=FALSE, fig.align = "center", fig.height = 7, fig.width = 10 }
DotPlot(seurat_object, features = rownames(seurat_object)[grep("S100A1", rownames(seurat_object))], group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
```

# *IL-14 (TXLNA)* and *IL-29 (IFNL1)*

## Which cells express *IL-14 (TXLNA)* and *IL-29 (IFNL1)*?

```{r, echo=FALSE, fig.align = "center", fig.height = 5, fig.width = 10 }
genes.bennet <- c("TXLNA", "IFNL1")
Idents(seurat_object) <- cell_type

                        FeaturePlot(
                            seurat_object, 
                            features = genes.bennet, 
                            order = TRUE, 
                            label = TRUE, 
                            repel = TRUE, 
                            min.cutoff = "q1", 
                            max.cutoff = "q99"
                        ) 
```

```{r, echo=FALSE, fig.align = "center", fig.height = 5, fig.width = 8 }
DotPlot(seurat_object, features = genes.bennet, group.by = cell_type, dot.scale = 6) +
  scale_color_gradient2(high = "red", low = "blue") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 10)
  )

```

# Marker of *TXLNA*-UP and *IFNL1*-UP cells {.tabset}

## *TXLNA*-UP cells

```{r, echo=FALSE, fig.align = "center", fig.height = 12, fig.width = 10}
TXLNA_UP <- mean(GetAssayData(object = seurat_object, slot = "data")["TXLNA",])
sd_TXLNA_UP <- sd(GetAssayData(object = seurat_object, slot = "data")["TXLNA",])
obj_TXLNA <- subset(x = seurat_object, subset = TXLNA > (TXLNA_UP+sd_TXLNA_UP))
number_TXLNA_up <- ncol(obj_TXLNA)
```

Number *TXLNA*-UP cells: `r number_TXLNA_up`

```{r, echo=FALSE, fig.align = "center", fig.height = 12, fig.width = 10}
if(number_TXLNA_up > 3){
  
  seurat_object$TXLNA_expr <- "-"
  seurat_object$TXLNA_expr[ colnames(obj_TXLNA)] <- "up"
  Idents(seurat_object) <- "TXLNA_expr"
  TXLNA_cells_markers <- FindMarkers(object = seurat_object, ident.1 = "up")
  save(TXLNA_cells_markers, file = here(output_data_path,"TXLNA-cells-markers.RData"))
  
  head(TXLNA_cells_markers, 10)
  
  TXLNA_coexpr <- head(rownames(TXLNA_cells_markers), 7)[-1]
  FeaturePlot(
      seurat_object,
      features = TXLNA_coexpr,
      order = TRUE,
      label = TRUE,
      repel = TRUE,
      min.cutoff = "q1",
      max.cutoff = "q99"
  )

}
```

## *IFNL1*-UP cells

```{r, echo=FALSE, fig.align = "center", fig.height = 12, fig.width = 10}
IFNL1_UP <- mean(GetAssayData(object = seurat_object, slot = "data")["IFNL1",])
sd_IFNL1_UP <- sd(GetAssayData(object = seurat_object, slot = "data")["IFNL1",])
obj_IFNL1 <- subset(x = seurat_object, subset = IFNL1 > (IFNL1_UP+sd_IFNL1_UP))
number_IFNL1_up <- ncol(obj_IFNL1)
```

Number *IFNL1*-UP cells: `r number_IFNL1_up`

```{r, echo=FALSE, fig.align = "center", fig.height = 12, fig.width = 10 }
if(number_IFNL1_up > 3){
  
  seurat_object$IFNL1_expr <- "-"
  seurat_object$IFNL1_expr[ colnames(obj_IFNL1)] <- "up"
  Idents(seurat_object) <- "IFNL1_expr"
  IFNL1_cells_markers <- FindMarkers(object = seurat_object, ident.1 = "up")
  save(IFNL1_cells_markers, file = here(output_data_path,"IFNL1-cells-markers.RData"))
  head(IFNL1_cells_markers, 10)
  IFNL1_coexpr <- head(rownames(IFNL1_cells_markers), 7)[-1]
  
  FeaturePlot(
      seurat_object,
      features = IFNL1_coexpr,
      order = TRUE,
      label = TRUE,
      repel = TRUE,
      min.cutoff = "q1",
      max.cutoff = "q99"
  )
  
}
```

# Annotation of Th1 and Th2 subpopulations {.tabset}

```{r, echo=FALSE, fig.align = "center", fig.height = 5, fig.width = 10}
th_cells <- colnames(seurat_object)[which(seurat_object@meta.data[, cell_type] %in% c("Th", "Tc", "Treg"))]
tcells_obj <- subset(seurat_object, cells = th_cells)

p1 <- DimPlot(tcells_obj, group.by = cell_type) +
  ggtitle("Celltypes")
p2 <-DimPlot(tcells_obj, group.by = "seurat_clusters")+
  ggtitle("Seurat clusters")
p3 <-FeaturePlot(
  tcells_obj,
  features = 'percent.mt') +
  ggtitle("Mitochondrial expression")

pw <- p1 + p2 
pw + plot_annotation(
  title = 'T-cells UMAP subset',
  subtitle = 'Subset of the dataset with only Tc, Th and Treg'
)  
```

## Th1-marker

https://www.rndsystems.com/resources/cell-markers/immune-cells/helper-t-cells/th1-cell-markers

```{r, echo=FALSE, fig.align = "center", fig.height = 20, fig.width = 10}
th1_marker <- intersect(c("STAT1", "STAT2", "IFNG", "IL2", "TNF", "CCR1", "CCR5", "CXCR3", "IL12RB2", "IL18R1"), rownames(seurat_object))
FeaturePlot(
  tcells_obj,
  features = th1_marker,
  ncol = 2,
  order = TRUE
  )
```

## Th2-marker

https://www.rndsystems.com/resources/cell-markers/immune-cells/helper-t-cells/th2-cell-markers

```{r, echo=FALSE, fig.align = "center", fig.height = 30, fig.width = 10}
th2_marker <- intersect(c("IL4", "IL5", "IL9", "IL10", "IL13", "IL21", "GATA3", "IRF4", "STAT5A", "STAT5B","STAT6", "CCR3", "CCR4", "CCR8", "CXCR4", "IL17RB", "IL4R", "IL33", "TSLP"), rownames(seurat_object))
FeaturePlot(
  tcells_obj, 
  features = th2_marker,
  ncol = 2,
  order = TRUE
  )

```

# Geneset Expression of GS th1-marker and th2-marker

```{r, echo=FALSE, fig.align = "center", fig.height=5, fig.width = 10}
th1_marker_list <- list(th1_marker)
tcells_obj <- AddModuleScore(
  object = tcells_obj,
  features = th1_marker_list,
  ctrl = 5,
  name = 'th1_marker'
)

th2_marker_list <- list(th2_marker)
tcells_obj <- AddModuleScore(
  object = tcells_obj,
  features = th2_marker_list,
  ctrl = 5,
  name = 'th2_marker'
)

pw <- FeaturePlot(
  tcells_obj,
  features = c('th1_marker1', 'th2_marker1'),
  ncol = 2,
  order = TRUE, 
  min.cutoff = "q1",
  max.cutoff = "q99"
  )

pw
```
